---
title: "sPLS performance"
output: html_document
date: "2025-07-30"
---



# Introduction

This page presents an application of the sPLS performance assessment. The sPLS method is a quite particular method : there are several predictions according to the components number selected in the model. The goal is almost to choose the best number of component in PLS regression in order to compute the best possible predictions. For that, we will use three datasets: 

- one is a dataset with only one response variable $Y$.

- the other is a dataset with four response variables $Y = (Y1,Y2,Y3,Y4)$.

- the last dataset contains real data about NIR spectra.

To access to predefined functions from sgPLSdevelop package and manipulate these datasets, run these lines :

```{r data, message=FALSE, warning=FALSE}
library(sgPLSdevelop)
library(pls)

data1 <- data.create(p = 10, list = TRUE)
data2 <- data.create(p = 10, q = 4, list = TRUE)
data(yarn)
data3 <- yarn
```
```{r, echo=FALSE}
print(paste("First dataset dimensions :",nrow(data1$D),"x",ncol(data1$D)))
print(paste("Second dataset dimensions :",nrow(data2$D),"x",ncol(data2$D)))
print(paste("Yarn dataset dimensions :",nrow(data3),"x",ncol(data3)))
```


For the two first datasets, the population is set to $n = 40$ by default, which is close to actual conditions. 
Let's also notice that, on average, the response $Y$ is a linear combination from the predictors $X$. Indeed, the function includes a matrix product $Y = XB + E$ with $B$ the weight matrix and $E$ matrix the gaussian noise. This linearity condition is important in order to have a good performance of the model, the PLS method using linearity combinaison. 

Now, it's time to train a PLS model for each dataset built or imported.

```{r pls, message=FALSE, warning=FALSE}
ncomp.max <- 8

# First model
X <- data1$X
Y <- data1$Y
model1 <- PLS(X,Y,mode = "regression", ncomp = ncomp.max)

# Second model
X <- data2$X
Y <- data2$Y
model2 <- PLS(X,Y,mode = "regression", ncomp = ncomp.max)

# Third model
X <- data3$NIR
Y <- data3$density
model3 <- PLS(X,Y,mode = "regression", ncomp = ncomp.max)
```


# PLS performance assessment using MSEP

An other way to assess such a model performance consists by using $MSEP$ criterion. $MSEP$ is computed as follow :

$MSEP = \frac{1}{nq} \sum_{i=1}^{n} \sum_{j=1}^{q} (Y_{i,j} - \hat{Y}_{i,j})^2$

### First model MSEP

```{r}
#perf.res1 <- perf.sPLS(model1)
#h.best <- perf.res1$h.best
#msep.best <- perf.res1$MSEP[h.best]
h.best <- 0
```
The `perf.PLS` gives us a optimal components number equal to $H =$ `r h.best`, therefore we suggest to select `r h.best` components in our first model.

### Second model MSEP

```{r}
#perf.res2 <- perf.sPLS(model2)
#h.best <- perf.res2$h.best
#msep.best <- perf.res2$MSEP[h.best]
```

The `perf.PLS` gives us a optimal components number equal to $H =$ `r h.best`, therefore we suggest to select `r h.best` components in our second model.

### Third model MSEP

```{r}
#perf.res3 <- perf.PLS(model3)
#h.best <- perf.res3$h.best
#msep.best <- perf.res3$MSEP[h.best]
```

The `perf.PLS` gives us a optimal components number equal to $H =$ `r h.best`, therefore we suggest to select `r h.best` components in our third model.